---
title: "Non-wear detection from ActiGraph data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-wear detection from ActiGraph data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      collapse = TRUE, comment = "#>", out.width = "75%",
                      fig.asp = 1 / 1.6, fig.width = 5, dpi = 600,
                      fig.retina = NULL)
pckg <- c("actigraph.sleepr", "dplyr", "lubridate", "ggplot2")
inst <- suppressMessages(lapply(pckg, library, character.only = TRUE))
theme_set(theme_light())
```

The `actigraph.sleepr` package implements two non-wear detection algorithms: Troiano (Troiano et al. 2008) and Choi (Choi et al. 2011). For illustration, let's use GT3X+ / ActiSleep+ sample data taken from ActiGraph's [online documentation](https://actigraph.desk.com). See the [sleep vignette](detect-sleep.html) for details on reading AGD files and collapsing epochs.

```{r }
file_10s <- system.file("extdata", "ActiSleepPlus-RawData-Day01-10sec.agd",
                        package = "actigraph.sleepr")
agdb_10s <- read_agd(file_10s)
agdb_60s <- collapse_epochs(agdb_10s, 60) %>%
  mutate(date = lubridate::as_date(timestamp)) %>%
  select(date, timestamp, starts_with("axis")) %>%
  group_by(date)
agdb_60s
```

Long stretches that consist almost entirely of zero counts (zero epochs) suggest that the device wasn't worn at all and therefore should be excluded from downstream analysis.

### Non-wear detection with the Troiano algorithm

The Troiano algorithm for detecting periods of non-wear formalizes a technique used to analyze the 2003-2004 NHANES data; the original SAS source code can be found at http://riskfactor.cancer.gov/tools/nhanes_pam/. The method has some flexibility as a non-wear period can contain a few nonzero epochs of artifactual movement (spikes).

activity_threshold
 : Highest activity level to be considered "zero"; an epoch with activity exceeding the threshold is considered a "spike". The default threshold is 0.

use_magnitude
 : If true, the magnitude of the vector (axis1, axis2, axis3) is used to measure activity; otherwise the axis1 value is used. The default is FALSE.

min_period_len
 : Minimum number of consecutive zero epoch to start a non-wear period. The default is 60.

max_nonzero_count
 : Epochs with activity greater than `max_nonzero_count` are labeled "zero". The default is `Inf`.

spike_tolerance
 : Also known as artifactual movement interval. At most `spike_tolerance` "nonzero" epochs can occur in sequence during a non-wear period without interrupting it. The default is 2.

spike_stoplevel
 : An activity spike that exceeds `spike_stoplevel` counts ends a non-wear period, even if the spike tolerance has not been reached. The default is 100.

The Troiano algorithm specifies that a non-wear period starts with `min_length` consecutive epochs/minutes of zero activity and ends with more than `spike_tolerance` epochs/minutes of nonzero activity.

```{r }
periods_nonwear <- apply_troiano(agdb_60s, min_period_len = 45)
periods_nonwear
```

Once non-wear periods are detected, we can further screen those intervals. For example, we can ignore non-wear periods that are too short.

```{r }
periods_nonwear %>% filter(length >= 60)
```

Or we can flag 24-hour day as invalid if the device was worn only for a short period of time.

```{r }
# Find the periods during which the device was worn
periods_wear <- complement_periods(periods_nonwear, agdb_60s,
                                   start_var = "period_start", end_var = "period_end")
periods_wear

# Each epoch either falls in a wear period (with period_id = 1, 2, 3, etc.)
# or in a non-wear period with (period_id = NA)
agdb_wear <- combine_epochs_periods(agdb_60s, periods_wear,
                                    start_var = "period_start", end_var = "period_end")
agdb_wear %>%
  filter(!is.na(period_id)) %>%
  summarise(time_worn = n())
```

### Non-wear detection with the Choi algorithm

The Choi algorithm extends the Troiano algorithm by requiring that short spikes of artifactual movement during a non-wear period are preceded and followed by \code{min_window_len} consecutive zero epochs.

```{r }
periods_nonwear <- apply_choi(agdb_60s, min_period_len = 45, min_window_len = 10)
periods_nonwear
```

### References

Troiano, Richard P, David Berrigan, Kevin W Dodd, Louise C Mâsse, Timothy Tilert, and Margaret McDowell. 2008. “Physical Activity in the United States Measured by Accelerometer.” *Medicine & Science in Sports & Exercise* 40 (1): 181–88.

Choi, Leena, Zhouwen Liu, Charles E. Matthews, and Maciej S. Buchowski. 2011. “Validation of Accelerometer Wear and Nonwear Time Classification Algorithm.” *Medicine & Science in Sports & Exercise* 43 (2): 357–64.
